The Poisson approximate likelihood compared to the particle filter

### 2024-07-27 ms.Rnw
addressing the email issues, correction on AIC

RUN_LEVEL = 1
NP_MIF       = switch(RUN_LEVEL, 10, 50000)
NMIF         = switch(RUN_LEVEL, 4,    100)
ncores       = switch(RUN_LEVEL, 2,    36)
top_n_fits   = switch(RUN_LEVEL, 1,    12)
sim_rep      = switch(RUN_LEVEL, 1,    100) # Total number of simulations

@@

<<OvOv-model, echo=F, warning=F, message=F>>=
# measurement model 
dmeas <- Csnippet("
                  if (ISNA(cases1)) {
                  lik = (give_log) ? 0 : 1;
                  } else {
                        lik =  dbinom(cases1, H1, q1, 1) +
                        dbinom(cases2, H2, q2, 1) +
                        dbinom(cases3, H3, q3, 1);

                    lik = (give_log) ? lik : exp(lik);

                    }")
rmeas <-  Csnippet("
                    cases1 = rbinom(H1, q1);
                    cases2 = rbinom(H2, q2);
                    cases3 = rbinom(H3, q3);
                  ")

@@

# define parameters (without betas)
params_fixed <- c(gamma=1, delta1=1/(5*52),delta2=1/(55*52), alpha=1/(78.86912*52), 
                  mu=0, N=82372825, omega=1/(1*52))
# full-estimate-rinit
rinit <- Csnippet("
    double m = N/(S10+I10+R10+S20+I20+R20+S30+I30+R30);
    I1=nearbyint(m*I10);
    I2=nearbyint(m*I20);
    I3=nearbyint(m*I30);
    S1=nearbyint(m*S10);
    S2=nearbyint(m*S20);
    S3=nearbyint(m*S30);
    R1=nearbyint(m*R10);
    R2=nearbyint(m*R20);
    R3=nearbyint(m*R30);
    H1 = 0;
    H2 = 0;
    H3 = 0;
")

@@

# Set to MLE
wwr_mle <- params_fixed

wwr_mle <- c(params_fixed, 
             "beta1" = 11.48,
             "beta2" = 0.25,
             "beta3" = 0.35,
             "phi" = 0.14,
             "beta11" = 0.16,
             "sigma_q" = 0.021,
             "sigma_xi" = 66.89,
             "S10" = 4.706102e-02,
             "S20" = 6.936707e-01,
             "S30" = 2.376236e-01,
             "I10" = 3.684589e-04,
             "I20" = 1.057388e-05,
             "I30" = 3.095681e-05,
             "R10" = 1.596669e-02,
             "R20" = 3.676601e-03,
             "R30" = 1.591447e-03)

wwr_mle_raw <- c(params_fixed, 
             "beta1" = 11.48,
             "beta2" = 0.25,
             "beta3" = 0.35,
             "phi" = 0.14,
             "beta11" = 0.16,
             "sigma_q" = 0.021,
             "sigma_xi" = 66.89)

pt <- pomp::parameter_trans(
  log = c("beta1","beta2","beta3","sigma_q","sigma_xi"),
  logit=c("beta11"),
  barycentric=c("S10","I10","R10",
                "S20","I20","R20",
                "S30","I30","R30"),
  toEst= pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
  fromEst= pomp::Csnippet("phi = M_2PI*expit(T_phi);")
)

read.table("real_rotavirus_metadata.txt") %>%
  rbind(data.frame(time=0,cases1=NA,cases2=NA,cases3=NA)) %>%
  arrange(time) -> dat

pomp(data = dat,
     times="time",
     t0=0,
     dmeasure = dmeas,
     rmeasure = rmeas,
     rprocess = discrete_time(step.fun = rproc, delta.t = 1/4),
     statenames = c("S1", "I1", "R1", "H1", 
                    "S2", "I2", "R2", "H2",
                    "S3", "I3", "R3", "H3", 
                    "q1", "q2", "q3"),
     paramnames = names(wwr_mle),
     accumvars=c("H1", "H2", "H3"),
     rinit=rinit,
     partrans = pt,
     params = wwr_mle
) -> sir
@

@@

<<maximization1, echo=F, warning=F, message=F>>=
set.seed(123)
sir_box <- list(
  beta1=c(10,15),
  beta2=c(0.2,0.4),
  beta3=c(0.3,0.5),
  phi=c(0.01,0.3),
  beta11=c(0.1,0.2),
  sigma_q=c(0.001,0.1),
  sigma_xi=c(65,70),
  S10 = c(0.01,0.05),
  I10 = c(0.0001, 0.0005),
  R10 = c(0.01, 0.05),
  S20 = c(0.01,0.7),
  I20 = c(0.000001,0.00002),
  R20 = c(0.001,0.005),
  S30 = c(0.01,0.5),
  I30 = c(0.000001, 0.0001),
  R30 = c(0.0001,0.005),
  #fixed params
  gamma = c(1,1),
  delta1=c(1/(5*52),1/(5*52)),
  delta2=c(1/(55*52),1/(55*52)),
  alpha=c(1/(78.86912*52),1/(78.86912*52)),
  mu=c(0,0),
  N=c(82372825,82372825),
  omega=c(1/(1*52),1/(1*52))
)

@@

# c(apply(sir_box,1,function(x)runif(1,min=x[1],max=x[2])),
#   params_fixed) -> starting

bake(file = paste0(out_dir, "mif2_maximization/round_01/ovov_mif_01.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_01 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01,beta2=0.01,beta3=0.01,
                    beta11=0.01,phi=0.01,sigma_q=0.01,sigma_xi=0.01,
                    S10=ivp(0.24),I10=ivp(0.24),R10=ivp(0.24),
                    S20=ivp(0.24),I20=ivp(0.24),R20=ivp(0.24),
                    S30=ivp(0.24),I30=ivp(0.24),R30=ivp(0.24)),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = design_matrix[i, ]
    ) 
  }
  mifs_global_01
})

bake(file = paste0(out_dir, "mif2_maximization/round_01/ovov_el_01.rds"),{
  el_01 <- eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
  el_01
})
@

@@

el_01 <- readRDS(paste0(out_dir,"mif2_maximization/round_01/ovov_el_01.rds"))
starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)

bake(file = paste0(out_dir,"mif2_maximization/round_02/ovov_mif_02.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_02 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
                    beta11=0.01*sd_02,phi=0.01*sd_02,
                    sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
                    S10=ivp(0.24)*sd_02,I10=ivp(0.24)*sd_02,R10=ivp(0.24)*sd_02,
                    S20=ivp(0.24)*sd_02,I20=ivp(0.24)*sd_02,R20=ivp(0.24)*sd_02,
                    S30=ivp(0.24)*sd_02,I30=ivp(0.24)*sd_02,R30=ivp(0.24)*sd_02),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting_values_02[[i]]
    ) 
  }
  mifs_global_02
})

bake(file = paste0(out_dir, "mif2_maximization/round_02/ovov_el_02.rds"),{
  el_02 <- eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
  el_02
})
@

@@

el_02 <- readRDS(paste0(out_dir, "mif2_maximization/round_02/ovov_el_02.rds"))
starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)

bake(file = paste0(out_dir, "mif2_maximization/round_03/ovov_mif_03.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_03 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
                    beta11=0.01*sd_03,phi=0.01*sd_03,
                    sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
                    S10=ivp(0.24)*sd_03,I10=ivp(0.24)*sd_03,R10=ivp(0.24)*sd_03,
                    S20=ivp(0.24)*sd_03,I20=ivp(0.24)*sd_03,R20=ivp(0.24)*sd_03,
                    S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting_values_03[[i]]
    ) 
  }
  mifs_global_03
})

bake(file = paste0(out_dir,"mif2_maximization/round_03/ovov_el_03.rds"),{
  el_03 <- eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
  el_03
})
@

<<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
mif_el <- readRDS(paste0(out_dir,"mif2_maximization/round_03/ovov_el_03.rds")) 
# Can be changed if more rounds needed
est_mle_vec <- mif_el[which.max(mif_el$logLik),]
@

@@

<<pf-sim-100, echo=F, warning=F, message=F>>=
# Set WWR's MLE
wwr_mle_raw <- c(params_fixed, 
             "beta1" = 11.48,
             "beta2" = 0.25,
             "beta3" = 0.35,
             "phi" = 0.14,
             "beta11" = 0.16,
             "sigma_q" = 0.021,
             "sigma_xi" = 66.89)

i <- 1

bake(file=paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"),{
  ovov_sim_100 <- list()
  registerDoFuture()
  plan(multicore, workers=ncores)

  repeat {
    # Simulation
    sim1 <- simulate(t0 = 0,
                     times = c(1:416),
                     dmeasure = dmeas,
                     rmeasure = rmeas,
                     rprocess = discrete_time(rproc, delta.t = 1/4),
                     statenames = c("S1", "I1", "R1", "H1", 
                                    "S2", "I2", "R2", "H2",
                                    "S3", "I3", "R3", "H3",
                                    "q1", "q2", "q3"),
                     obsnames = c("cases1", "cases2", "cases3"),
                     paramnames = names(wwr_mle_raw),
                     accumvars = c("H1", "H2", "H3"),
                     rinit = rinit_wwr,
                     params = wwr_mle_raw)

    sim_data <- as.data.frame(sim1)
    if (!any(sim_data == 0)) {
      ovov_sim_100[[length(ovov_sim_100) + 1]] <- sim_data[, c("cases1", "cases2", "cases3")]
      i <- i + 1  # Increment only if the condition is satisfied
    } else {
      print(paste("Condition not met for iteration", i, "- retrying"))
    }

    if (length(ovov_sim_100) >= sim_rep) break  # Exit the loop when the desired number of successful iterations is reached
  }
  ovov_sim_100
})

# load("ovov_sim_100.Rdata")
@

<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"))
bake(file=paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
  sir_wwr <- list()
  pfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)

  for(i in 1:length(ovov_sim_100)){
  data.frame(time=c(1:416), ovov_sim_100[[i]]) |>  pomp(t0=0,
           time="time",
           rprocess=discrete_time(rproc,delta.t=1/4),
           rinit=rinit_wwr,
           dmeasure=dmeas,
           rmeasure=rmeas,
           accumvars=c("H1","H2","H3"),
           statenames=c("S1","S2","S3",
                        "I1","I2","I3",
                        "R1","R2","R3",
                        "H1","H2","H3",
                        "q1","q2","q3"),
           paramnames = names(wwr_mle_raw)
      ) -> sir_wwr[[i]]

    foreach(j=1:ncores, .combine=c,
            .options.future=list(seed=998468235L)
    ) %dofuture% {
      pfilter(sir_wwr[[i]],Np=NP_MIF,save.states=T, params=wwr_mle_raw)
    } -> pfs

@@

init_dist <- c(3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092)
# initial_guess <- c(runif(1,5,20),runif(1,0.05,0.5),runif(1,0.2,1),runif(1,0.05,0.4),runif(1,0.05,0.4),runif(1,0.05,0.2),runif(1,10,100))
initial_guess <- c(11.48,0.25,0.35,0.14,0.16,0.021,66.89)

### Just a filter
Sys.time()

bake(file=paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_sim_100.rds"),{
  palfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)
  for(i in 1:length(ovov_sim_100)){
    pal_sim_dat <- ovov_sim_100[[i]] |> as.matrix()
    lik_list <- rotavirus_SMC_qropxi(init_dist,pal_sim_dat , 9, 
                                     c(initial_guess[1],initial_guess[2],initial_guess[3],
                                       initial_guess[4],initial_guess[5]), 
                                     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
                                     norm_par = c(0.07,initial_guess[6]), 
                                     t(prop), NP_MIF,1)
    palfilter_ovov_sim_100[i] <- lik_list$log_lik
  }
  palfilter_ovov_sim_100
})

Sys.time()

# lik_list$ll_storage |> sum()

# x <- c()
# for(i in 1:10){
#   x[i] <- any(ovov_sim_100[[i]]==0)
# }
@

<<plot1,echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.The particle filter and PAL used 50,000 particles.The red line is the line $Lik(pfilter)=Lik(palfilter)$.On average, the particle filter likelihood is $7.7$ log units higher than the PAL. {This may not be wrong, since the likelihood evaluated by pfilter is not fully optimized by employing many particles.}'>>=
data <- data.frame(pfilter = pfilter_ovov_sim_100, palfilter = palfilter_ovov_sim_100)

@@

<<pfilter_org, echo=FALSE, message=FALSE, warning=FALSE>>=
dat[-1, ] |>
  pomp(t0=0,
       time="time",
       rprocess=discrete_time(rproc,delta.t=1/4), 
       rinit=rinit_wwr,
       dmeasure=dmeas,
       rmeasure=rmeas,
       accumvars=c("H1","H2","H3"),
       statenames=c("S1","S2","S3",
                    "I1","I2","I3",
                    "R1","R2","R3",
                    "H1","H2","H3",
                    "q1","q2","q3"),
       paramnames = names(wwr_mle_raw)
  ) -> m1

bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_org.rds"),{
  registerDoFuture()
  plan(multicore, workers = ncores)
  foreach(i=1:ncores, .combine=c,
          .options.future=list(seed=998468235L)
  ) %dofuture% {
    pfilter(m1,Np=NP_MIF,params=wwr_mle_raw)
  } -> pfs
  pfs
})

bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_list_cond_lik.rds"),{
  list_cond_lik <- list()
  registerDoFuture()
  plan(multicore, workers = ncores)
  for(k in 1:ncores){
    list_cond_lik[[k]] <- pfs[[k]]@cond.logLik
  }
  list_cond_lik
})
@

<<palfilter_org, echo=FALSE>>=
load("real_rotavirus_metadata.Rdata")
prop <- 420

init_dist <- c(3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092)
# initial_guess <- c(runif(1,5,20),runif(1,0.05,0.5),runif(1,0.2,1),runif(1,0.05,0.4),runif(1,0.05,0.4),runif(1,0.05,0.2),runif(1,10,100))
initial_guess <- c(11.48,0.25,0.35,0.14,0.16,0.021,66.89)

@@

Sys.time()
bake(file=paste0(out_dir,"pfilter/original_dataset/palfilter_org.rds"),{
  realdat <- ovov_sim_100[[i]] |> as.matrix()
  pal_lik_list_org <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
                                   c(initial_guess[1],initial_guess[2],
                                     initial_guess[3],initial_guess[4],initial_guess[5]), 
                                   gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
                                   norm_par = c(0.07,initial_guess[6]), 
                                   t(prop), NP_MIF,1)
  pal_lik_list_org
})
Sys.time()
@

<<plot2, echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances. The blue line represents overlapping results from 36 SMC computations, while the red line is derived from PAL. The main sources of likelihood shortfall of SMC are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=
# Create your data
time <- 1:416
ncores <- length(list_cond_lik)

# Combine the data into a single dataframe for SMC
data_list <- lapply(1:ncores, function(i) {
  data.frame(time = time, cond_loglik = list_cond_lik[[i]], method = "SMC")
})

@@

# Create the PAL data frame
df_pal <- data.frame(time = time, cond_loglik = pal_lik_list_org$ll_storage, method = "PAL")

@@

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering % Centers the table
\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} % This is effectively the table title
\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
\begin{tabular}{|c c c|} 
 \hline
  & PAL & SMC  \\
  \hline
 OvOv & \Sexpr{pal_lik_list_org$log_lik} & $\Sexpr{logmeanexp(sapply(pfs, logLik))}$ \\ 
 \hline
\end{tabular}
\end{table}

>>>>

RUN_LEVEL = 1
NP_MIF       = switch(RUN_LEVEL, 10, 50000)
NMIF         = switch(RUN_LEVEL, 5,    100)
ncores       = switch(RUN_LEVEL, 4,    36)
top_n_fits   = switch(RUN_LEVEL, 1,    12)
sim_rep      = switch(RUN_LEVEL, 5,    100) # Total number of simulations

@@

<<OvOv-model, echo=F, warning=F, message=F>>=
# measurement model 
dmeas <- Csnippet("
  lik =  dbinom(cases1, H1, q1, 1) +
  dbinom(cases2, H2, q2, 1) +
  dbinom(cases3, H3, q3, 1);"
)
rmeas <-  Csnippet("
  cases1 = rbinom(H1, q1);
  cases2 = rbinom(H2, q2);
  cases3 = rbinom(H3, q3);
")

@@

# full-estimate-rinit
rinit_full <- Csnippet("
  double m = N/(S10+I10+R10+S20+I20+R20+S30+I30+R30);
  I1=nearbyint(m*I10);
  I2=nearbyint(m*I20);
  I3=nearbyint(m*I30);
  S1=nearbyint(m*S10);
  S2=nearbyint(m*S20);
  S3=nearbyint(m*S30);
  R1=nearbyint(m*R10);
  R2=nearbyint(m*R20);
  R3=nearbyint(m*R30);
  H1 = 0;
  H2 = 0;
  H3 = 0;
")

@@

# define parameters (without betas)
params_fixed <- c(gamma=1, delta1=1/(5*52),delta2=1/(55*52), alpha=1/(78.86912*52), 
                  mu=0, N=82372825, omega=1/(1*52))
# Set to MLE
params_wwr <- c(params_fixed, 
  "beta1" = 11.48,
  "beta2" = 0.25,
  "beta3" = 0.35,
  "phi" = 0.14,
  "beta11" = 0.16,
  "sigma_q" = 0.021,
  "sigma_xi" = 66.89)

params_full <- c(params_wwr,
  "S10" = 4.706102e-02,
  "S20" = 6.936707e-01,
  "S30" = 2.376236e-01,
  "I10" = 3.684589e-04,
  "I20" = 1.057388e-05,
  "I30" = 3.095681e-05,
  "R10" = 1.596669e-02,
  "R20" = 3.676601e-03,
  "R30" = 1.591447e-03)

pt_full <- pomp::parameter_trans(
  log = c("beta1","beta2","beta3","sigma_q","sigma_xi"),
  logit=c("beta11"),
  barycentric=c("S10","I10","R10",
                "S20","I20","R20",
                "S30","I30","R30"),
  toEst= pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
  fromEst= pomp::Csnippet("phi = M_2PI*expit(T_phi);")
)

pt_wwr <- pomp::parameter_trans(
  log = c("beta1","beta2","beta3","sigma_q","sigma_xi"),
  logit=c("beta11"),
  toEst= pomp::Csnippet("T_phi = logit(phi/(M_2PI));"),
  fromEst= pomp::Csnippet("phi = M_2PI*expit(T_phi);")
)

read.table("real_rotavirus_metadata.txt") %>%
  arrange(time) -> dat

statenames <- c(
  "S1", "I1", "R1", "H1",
  "S2", "I2", "R2", "H2",
  "S3", "I3", "R3", "H3",
  "q1", "q2", "q3"
)

sir_wwr <- pomp(data = dat,
  times="time",
  t0=0,
  dmeasure = dmeas,
  rmeasure = rmeas,
  rprocess = discrete_time(step.fun = rproc, delta.t = 1/4),
  statenames = statenames,
  paramnames = names(params_wwr),
  accumvars= c("H1", "H2", "H3"),
  rinit= rinit_wwr,
  partrans = pt_wwr,
  params = params_wwr
)

sir_full <- pomp(sir_wwr,
  rinit = rinit_full,
  params = params_full,
  partrans = pt_full,
  statenames = statenames,
  paramnames = names(params_full)
)
@

@@

<<maximization1, echo=F, warning=F, message=F>>=
set.seed(123)
sir_box <- list(
  beta1=c(10,15),
  beta2=c(0.2,0.4),
  beta3=c(0.3,0.5),
  phi=c(0.01,0.3),
  beta11=c(0.1,0.2),
  sigma_q=c(0.001,0.1),
  sigma_xi=c(65,70),
  S10 = c(0.01,0.05),
  I10 = c(0.0001, 0.0005),
  R10 = c(0.01, 0.05),
  S20 = c(0.01,0.7),
  I20 = c(0.000001,0.00002),
  R20 = c(0.001,0.005),
  S30 = c(0.01,0.5),
  I30 = c(0.000001, 0.0001),
  R30 = c(0.0001,0.005),
  gamma = c(1,1),
  delta1=c(1/(5*52),1/(5*52)),
  delta2=c(1/(55*52),1/(55*52)),
  alpha=c(1/(78.86912*52),1/(78.86912*52)),
  mu=c(0,0),
  N=c(82372825,82372825),
  omega=c(1/(1*52),1/(1*52))
)

@@

bake(file = paste0(out_dir, "mif2_maximization/round_01/mifs_global_01.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_01 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir_full,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01,beta2=0.01,beta3=0.01,
        beta11=0.01,phi=0.01,sigma_q=0.01,sigma_xi=0.01,
        S10=ivp(0.24),I10=ivp(0.24),R10=ivp(0.24),
        S20=ivp(0.24),I20=ivp(0.24),R20=ivp(0.24),
        S30=ivp(0.24),I30=ivp(0.24),R30=ivp(0.24)),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = design_matrix[i, ]
    ) 
  }
  mifs_global_01
})

bake(file = paste0(out_dir, "mif2_maximization/round_01/el_01.rds"),{
  el_01 <- eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
  el_01
})
@

@@

el_01 <- readRDS(paste0(out_dir,"mif2_maximization/round_01/el_01.rds"))
starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)

bake(file = paste0(out_dir,"mif2_maximization/round_02/mifs_global_02.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_02 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      mifs_global_01[[1]],
      rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
                beta11=0.01*sd_02,phi=0.01*sd_02,
                sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
                S10=ivp(0.24)*sd_02,I10=ivp(0.24)*sd_02,R10=ivp(0.24)*sd_02,
                S20=ivp(0.24)*sd_02,I20=ivp(0.24)*sd_02,R20=ivp(0.24)*sd_02,
                S30=ivp(0.24)*sd_02,I30=ivp(0.24)*sd_02,R30=ivp(0.24)*sd_02),
      params = starting_values_02[[i]]
    ) 
  }
  mifs_global_02
})

bake(file = paste0(out_dir, "mif2_maximization/round_02/el_02.rds"),{
  el_02 <- eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
  el_02
})
@

@@

el_02 <- readRDS(paste0(out_dir, "mif2_maximization/round_02/el_02.rds"))
starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)

bake(file = paste0(out_dir, "mif2_maximization/round_03/mifs_global_03.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_03 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      mifs_global_02[[1]],
      rw.sd = rw_sd(beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
                beta11=0.01*sd_03,phi=0.01*sd_03,
                sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
                S10=ivp(0.24)*sd_03,I10=ivp(0.24)*sd_03,R10=ivp(0.24)*sd_03,
                S20=ivp(0.24)*sd_03,I20=ivp(0.24)*sd_03,R20=ivp(0.24)*sd_03,
                S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03),
      params = starting_values_03[[i]]
    ) 
  }
  mifs_global_03
})

bake(file = paste0(out_dir,"mif2_maximization/round_03/el_03.rds"),{
  el_03 <- eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
  el_03
})
@

<<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
el_03 <- readRDS(paste0(out_dir,"mif2_maximization/round_03/el_03.rds")) 
# Can be changed if more rounds needed
est_mle_vec <- el_03[which.max(el_03$logLik),]
@

@@

<<sim-100, echo=F, warning=F, message=F>>=
bake(file=paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"),{
  ovov_sim_100 <- list()
  registerDoFuture()
  plan(multicore, workers=ncores)

  i <- 1
  repeat {
    # Simulation
    sim_data <- simulate(sir_wwr, format="data.frame") |>
      subset(select=c("time","cases1","cases2","cases3"))

    if (!any(sim_data == 0)) {
      ovov_sim_100[[length(ovov_sim_100) + 1]] <- sim_data
      i <- i + 1  
    } else {
      print(paste("Condition not met for iteration", i, "- retrying"))
    }

    if (length(ovov_sim_100) >= sim_rep) break  
  }
  ovov_sim_100
})
@

<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"))
bake(file=paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
  pfilter_obj <- list()
  pfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)

  for(i in 1:length(ovov_sim_100)){
    pomp(data = ovov_sim_100[[i]],
    times="time",
    t0=0,
    dmeasure = dmeas,
    rmeasure = rmeas,
    rprocess = discrete_time(step.fun = rproc, delta.t = 1/4),
    statenames = statenames,
    paramnames = names(params_wwr),
    accumvars= c("H1", "H2", "H3"),
    rinit= rinit_wwr,
    partrans = pt_wwr,
    params = params_wwr
    )  -> pfilter_obj[[i]]

    foreach(j=1:ncores, .combine=c,
            .options.future=list(seed=998468235L)
    ) %dofuture% {
      pfilter(pfilter_obj[[i]],Np=NP_MIF,save.states=T, params=params_wwr)
    } -> pfs

@@

init_dist <- c(3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092)
initial_guess <- c(11.48,0.25,0.35,0.14,0.16,0.021,66.89)

bake(file=paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_sim_100.rds"),{
  palfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)
  for(i in 1:length(ovov_sim_100)){
    pal_sim_dat <- ovov_sim_100[[i]][,-1] |> as.matrix()
    lik_list <- rotavirus_SMC_qropxi(init_dist,pal_sim_dat , 9, 
      c(initial_guess[1],initial_guess[2],initial_guess[3],initial_guess[4],initial_guess[5]), 
      gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
      norm_par = c(0.07,initial_guess[6]), 
      t(prop), NP_MIF,1)
    palfilter_ovov_sim_100[i] <- lik_list$log_lik
  }
  palfilter_ovov_sim_100
})

# lik_list$ll_storage |> sum()
@

<<plot1-100-sim-compare,echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Log-likelihoods computed using two filtering methods for 100 randomly simulated datasets at the MLE of the OvOv model.The particle filter and PAL used 50,000 particles.The red line is the line $Lik(pfilter)=Lik(palfilter)$.On average, the particle filter likelihood is $7.7$ log units higher than the PAL. {This may not be wrong, since the likelihood evaluated by pfilter is not fully optimized by employing many particles.}'>>=
data <- data.frame(pfilter = pfilter_ovov_sim_100, palfilter = palfilter_ovov_sim_100)

@@

<<pfilter_wwr, echo=FALSE, message=FALSE, warning=FALSE>>=
bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_wwr.rds"),{
  registerDoFuture()
  plan(multicore, workers = ncores)
  foreach(i=1:ncores, .combine=c,
          .options.future=list(seed=998468235L)
  ) %dofuture% {
    pfilter(sir_wwr,Np=NP_MIF,params=params_wwr)
  } -> pfilter_wwr
  pfilter_wwr
})

bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_wwr_cond_lik.rds"),{
  pfilter_wwr_cond_lik <- list()
  registerDoFuture()
  plan(multicore, workers = ncores)
  for(k in 1:ncores){
    pfilter_wwr_cond_lik[[k]] <- pfilter_wwr[[k]]@cond.logLik
  }
  pfilter_wwr_cond_lik
})
@

<<palfilter_wwr, echo=FALSE>>=
realdat <- as.matrix(dat[,-1])
prop <- 420

init_dist <- c(3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092)
initial_guess <- c(11.48,0.25,0.35,0.14,0.16,0.021,66.89)

@@

bake(file=paste0(out_dir,"pfilter/original_dataset/palfilter_lik_list_wwr.rds"),{
  palfilter_lik_list_wwr <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
    c(initial_guess[1],initial_guess[2],initial_guess[3],initial_guess[4],initial_guess[5]),     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
    norm_par = c(0.07,initial_guess[6]), 
    t(prop), NP_MIF,1)
  palfilter_lik_list_wwr
})
@

<<plot2-cond-loglik, echo=FALSE, fig.height=4, fig.width=5.5, fig.cap='Conditional log-likelihoods computed using two methods for the real rotavirus dataset, with the SMC method replicated 36 times due to its high variances. The blue line represents overlapping results from 36 SMC computations, while the red line is derived from PAL. The main sources of likelihood shortfall of SMC are at time points $t=1, 2, 3, 11, 81, 194$, and $325-333$.'>>=
# Create your data
time <- 1:416

# Combine the data into a single dataframe for SMC
data_list <- lapply(1:ncores, function(i) {
  data.frame(time = time, cond_loglik = pfilter_wwr_cond_lik[[i]], method = "SMC")
})

@@

# Create the PAL data frame
df_pal <- data.frame(time = time, cond_loglik = palfilter_lik_list_wwr$ll_storage, method = "PAL")

@@

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering % Centers the table
\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} % This is effectively the table title
\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
\begin{tabular}{|c c c|} 
 \hline
  & PAL & SMC  \\
  \hline
 OvOv & \Sexpr{palfilter_lik_list_wwr$log_lik} & $\Sexpr{logmeanexp(logLik(pfilter_wwr))}$ \\ 
 \hline
\end{tabular}
\end{table}
